---
title: "COVID19 Analysis"
author: "G.U"
date: "2024-02-29"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(stringr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
```

## COVID-19 Analysis:

Data Source: "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

Problem Statement: To showcase all the steps in the data science process, reproducibility,clear writing,validity of  analysis and identification of bias, using the COVID-19 Dataset from Johns Hopkins.

Data Description:            The data contains daily time series summary tables including confirmed, deaths and recovered within the US as well as globally.  


```{r data_load}
# Base URL for raw files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

# file urls
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)

#reading the data

global_cases <- read.csv(urls[1])
global_deaths <- read.csv(urls[2])
US_cases <- read.csv(urls[3])
US_deaths <- read.csv(urls[4])

#cleaning up global_cases and global_deaths

global_cases <- global_cases %>% pivot_longer(cols=
                -c('Province.State',
                   'Country.Region', 'Lat', 'Long'),
                  names_to = "date",
                  values_to = "cases") %>% 
                  select(-c(Lat, Long)) %>%
                  mutate(date = gsub("^X", "", date), # Remove the 'X' prefix
                  date = mdy(date)) # Parse the date

global_deaths <- global_deaths %>% pivot_longer(cols=
                -c('Province.State',
                   'Country.Region', 'Lat', 'Long'),
                  names_to = "date",
                  values_to = "deaths") %>% 
                  select(-c(Lat, Long)) %>%
                  mutate(date = gsub("^X", "", date), # Remove the 'X' prefix
                  date = mdy(date)) # Parse the date

#combining global_cases with global_deaths

global <- global_cases %>%
  full_join(global_deaths, by = c("Province.State", "Country.Region", "date")) %>%
  rename(Country_Region = 'Country.Region',
         Province_State = 'Province.State')

#filtering out cases = 0
global <- global %>% filter(cases > 0)

summary(global)

#cleaning US data
US_cases <- US_cases %>% pivot_longer(cols=-(UID:Combined_Key),
                  names_to = "date",
                  values_to = "cases") %>% 
                  select(Admin2:cases) %>%
                  select(-c(Lat, Long_)) %>%
                  mutate(date = gsub("^X", "", date), # Remove the 'X' prefix
                  date = mdy(date)) # Parse the date

US_deaths <- US_deaths %>% pivot_longer(cols=-(UID:Population),
                  names_to = "date",
                  values_to = "deaths") %>% 
                  select(Admin2:deaths) %>%
                  select(-c(Lat, Long_)) %>%
                  mutate(date = gsub("^X", "", date), # Remove the 'X' prefix
                  date = mdy(date)) # Parse the date

US <- US_cases %>% full_join(US_deaths, by = c("Admin2", "Province_State", "Country_Region", "Combined_Key", "date"))

US <- US %>% filter(cases > 0)


global <- global %>%
          unite("Combined_Key",
                c(Province_State, Country_Region),
                sep = ", ",
                na.rm = TRUE,
                remove = FALSE)


uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

#Adding Population in global
global <- global %>%
  left_join(uid, by = c("Province_State" = "Province_State", "Country_Region" = "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

```

## Including Plots for analysis

In this section I perform various exploratory data analysis on the data:

```{r exploratory_analysis, echo=FALSE}
#grouping US data

US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases, na.rm = TRUE), deaths = sum(deaths, na.rm = TRUE), 
            Population = sum(Population, na.rm = TRUE), .groups = "drop") %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

#grouping US totals
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases, na.rm = TRUE), deaths = sum(deaths, na.rm = TRUE), 
            Population = sum(Population, na.rm = TRUE), .groups = "drop") %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>% 
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

#visualizing the US totals
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y= deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in US", y= NULL)

#visualizing for New York
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y= deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ",state), y= NULL)

max(US_totals$date)

max(US_totals$deaths)

#Transforming US data further
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))

#visualizing new cases and new deaths
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y= new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in US", y= NULL)


#visualizing new cases and new deaths in New York
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y= new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ",state), y= NULL)


#looking for the best states
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000*cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

US_state_totals %>% 
  slice_min(deaths_per_thou, n = 10) %>%
select(deaths_per_thou, cases_per_thou, everything())

#looking for the worst states
US_state_totals %>% 
  slice_max(deaths_per_thou, n = 10) %>%
select(deaths_per_thou, cases_per_thou, everything())
```

## Creating a Model

In this section I created a linear model to show the relationship between deaths per thousand and cases per thousand.

BIAS: 

As I build this model and remember how intense the COVID season actually was, I can identify a few avenues for bias:
1. The dataset and model do not account for the demography of the deaths or reported cases. This means that the model does acknowledge the fact that (for example) older individuals who died during the period due to other health issues could have been grouped in this bucket due to similarity of symptoms. 

2. Also, an individual can view the chart of cases or deaths by state differently depending on how personally impacted they were. Although this may seem anecdotal, however, an individual could feel like a given state's numbers should be higher if they personally had relatives or friends in that state that were affected by COVID. 

```{r modelling, echo=FALSE}

#Developing a linear model

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)

#creating a new dataset with the predicted column
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

#visualizing the prediction
US_tot_w_pred %>% ggplot() +
geom_point(aes(x = cases_per_thou, y = deaths_per_thou),color = "blue")+
geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
